A sustainable approach to tourist signage on (historical) trails

### 2023-07-20 paper.tex
Sostituito storico compresso di Letizia (manca biblio)

% The village of Casoli is located on the Lima stream. To the south-east of the settlement lies a modest relief with a small glacial lake. The name of the hill ({\em "Tanette", in italian small lair} reveals the presence of karstic cavities, some inhabitated between the Paleolithic and the Iron Age . These frequentations can be attributed to a seasonal exploitation of resources and, later, with the climatic improvement, as stations on the trans-Apennine routes \cite{men76, gia96, pal63, zec72a, zec72b}.

South-east of the today location of Casoli lies a modest relief and a small glacial lake. The name of the hill ({\em "Tanette", in italian small lair} reveals the presence of karstic cavities, some inhabitated between the Paleolithic and the Iron Age and used also as stations on the trans-Apennine routes \cite{men76, gia96, pal63, zec72a, zec72b}.

@@

We have little evidence of settlements in the mountainous hinterland, and the scarce archaeological traces are concentrated in the cave of Buca La Piella, investigated in 1975 by the Centre for Archaeological Studies of Lucca; it has two entrances joined by a walkable tunnel and rooms of discrete dimensions that overlook the outside. In addition to numerous faunal remains and fragments of locally produced common pottery, twenty bronze coins belonging to the 3rd century AD, bronze and lead objects were found.

% Just outside the Casoli settlement, ceramic fragments datable between the 3rd century B.C. and the 1st-2nd century A.D. emerged from an excavation for the construction of a bowling green, testifying, together with other documented evidence in the area of Bagni di Lucca, how Ligurian populations settled in Val di Lima at least from the 3rd century B.C. onwards.

% In this period, the settlement must have been rather scattered and in small nuclei. The dedication of the Latin colony of Lucca in 180 B.C. marked a decisive turning point in the Romanisation of the area and within a few years there was a definitive subjugation of the Ligurian populations, accompanied by a rapid acculturation \cite{gia96, cia05}.

% From the data currently available, it can be stated that the mountainous hinterland retained a relatively marginal role. Archaeological traces for the Roman era are scarce, apart from the material that emerged from Buca La Piella, located east of Casoli along the Solco del Monte.

%The cave was discovered and investigated in 1975 by the Centre for Archaeological Studies of Lucca; it has two entrances joined by a walkable tunnel and rooms of discrete dimensions that overlook the outside. In addition to numerous faunal remains and fragments of locally produced common pottery, twenty bronze coins belonging to the 3rd century AD, bronze and lead objects were found.

% This confirms a rarefied or seasonal frequentation with a possible division in “fundi”, probably linked to the exploitation of the mountain's natural resources, the products of the forest and pasture, as well as having represented a connection route with the Po Valley area, through the Apennine passes \cite{gia96, men81, cia03}.

In addition to the scarcity of archaeological remains, there is a profound silence in the historical and literary sources for the Val di Lima from Roman and Late Antique times, although it can be assumed that the Val di Lima followed the fate of the city of Lucca, first with the occupation of the Goths and the brief Byzantine rule and then with the Longobard invasion and the formation of the Duchy of Lucca. In the hinterland, the Lombard settlement was concentrated according to a system consisting of castles and villages, mostly evidenced by documents from the 8th and 9th centuries and the remains of burial areas. In Longobard and Carolingian times, Lucca's Val di Lima was one of the three administrative districts into which the mountains were divided and was called “fines Contronenses” \cite{qui02, cia06, cia11}.

The only find from this period is the Grotta di Arzale, also known as the 'Grotta Murata', a rock shelter that opens up in the locality of Sevilucchio, north-west of Casoli along a historic road connecting Cocciglia to Casoli. Fictile artefacts, defined as being from the 'barbarian' era, have been found here; in reality, the bibliographic and oral sources collected have indicated fragments of an olla with threading, the attestation of which in the archaeological contexts of Lucca and northern Tuscany starts from the 6th-7th century and may reach as far as the end of the 11th century \cite{gia96}.

The first attestation of the settlements of Casoli and Lacu (Lake) (around the area of the Lake Casoli) dates back to the 10th century, the communities of which appear to belong to the parish of Vico Pancellorum. In this period, Bagni di Lucca was subdivided into four plebates, to which numerous villages were attached \cite{bar41, gia96}. The demographic and material entity of these agglomerates are rather elusive, but they testify to a wide-ranging distribution of the settlement with a strengthening of the social function of the churches as seats of aggregation of the population. Most documents show a widespread allivation by the bishop of Lucca of many properties, tithes, annuities and offerings of the parishes, a practice that incentivised the fractioning and alienation of ecclesiastical property in favour of certain members of the city aristocracy. In this way, they secured an accumulation of funds and power, heritages that were to the nucleus of the wealth of the future lords over the following centuries, until they emerged in
the territory by building their castles and constructing seigniorial domains centred on these areas \cite{qui02, gia96, for12, for15}.

Written sources mention a castle in Casoli from 1180, the year in which the fortification is in the hands of the Bishop of Lucca, but its foundation must be earlier, probably built during the 11th- 12th centuries. It occupies the summit of Colle delle Tanette, towering to the north of the village.

Between the 13th and 15th centuries, the fortification was at the centre, along with the other castles of the Val di Lima, of clashes between Lucca, Pisa and Florence, with alternating fortunes, as it was a strategic border area for the power of Lucca. Today, very little is preserved and the area has been partly inhabited. The fortified complex must have been equipped with
two walls, a tower on the highest portion of the hill that is poorly preserved and two buildings.

According to some oral sources, part of the castle stones were re-used to construct the surrounding buildings \cite{gia96, for12, rom16}. The same document from 1180 mentions the church of San Donato, located in the main square on the north-western edge of the village of Casoli. The current structure dates back to the 12th-13th centuries, with subsequent renovations and an enlargement of the naves that partially incorporated the base of the bell tower. The façade is surmounted by a plastered moulded tympanum, while the side aisles are adorned with moulded cornices. Initially dedicated only to San Donato, it acquired the double dedication after the abandonment of the church of Sant'Andrea de Lacu. The latter is located on a plateau to the east of Lake Casoli and is attested from 1260 in the Estimi of the diocese of Lucca as being subject to the parish of Vico Pancellorum \cite{ber18, gia96, cap17}, but already from the pastoral visit of 1358 it appears to have been united to the church of Casoli and from the pastoral visit of 1467 we learn of its state of abandonment \cite{gia96, con12}. This was probably due to the depopulation of the lake area and the simultaneous strengthening of the town of Casoli, which was fortified and better protected during this unstable phase.

Today, the romanesque church of Sant'Andrea de Lacu is in a state of decay, with a rectangular plan, ending on the east side with a semi-circular apse; the masonry is characterised by squared lithic blocks arranged in regular rows. Inside, there is a reused element of the previous building walled into the northern perimeter, testifying to an older origin, probably contemporary with the settlement of Lacu (Lake) mentioned in sources from the 10th century and no longer visible today.

Nearby are the remains of a quadrangular building with masonry in lithic blocks arranged in fairly regular rows, dating back to the Middle Ages (13th century), possibly a residential building.

In the early modern age, villages that survived the demographic crisis of the 14th-15th centuries experienced a progressive architectural renewal between the 16th and 18th centuries, which still largely characterizes the settlements today. Within Casoli, a series of residential buildings with imposing dated portals are preserved, some with courtyards, and mansions that denote a discrete deployment of resources by wealthy social classes. There is also a well-preserved bread oven, typically used by the entire community, which remained in use at least for village festivals until a few decades ago.

Two oratories in the village date back to the modern age (17th-18th centuries): one dedicated to the Madonna of Loreto, flanked by a wash-house, and the other called Madonna di Castello and characterised by an entrance enclosure, located along the cobbled road that retraces the ancient road linking the village with the summit area of the medieval castle \cite{gia96}.

Casoli was connected to the valley floor by two main roads attested in the Borbonic Cadastre, which from the village follow the gentler slopes to the east and south-west to ford the Lima torrent thanks to the Ponte Maggio and Ponte Nero bridges, still preserved today. Towards the hilly area, an articulated road network is evidenced, which is still legible in the landscape and in part is taken up by the CAI trail system. In particular, from the village starts the road that led toward Lucchio crossing the area of Casoli Lake. Along it are the chapel of Madonna di Col di Piano and two “metati” referable to the contemporary age, used for drying chestnuts. The entire area of the lake is covered with crops of centuries-old chestnut trees, typical of the Lima Valley, as of the entire Apennine area. Referred to as the "bread tree," chestnut fruits and the flour derived from them were a staple of the mountain diet until the mid-1900s, and today they are increasingly less common due to the disease that has affected their specimens and the slow replacement by allochthonous species \cite{buc92, puc10}.

@@

% TODO: \usepackage{graphicx} required
\begin{figure}
	\centering
	\includegraphics[width=\linewidth]{figure/mappa}
	\caption[Map with QR-tag positions]{}
	\label{fig:mappa}
\end{figure}

>>>>

 The hill village of Casoli is located south of the Lima stream. The name of the hill ({\em "Tanette", in italian small lair} reveals the presence of karstic cavities, some inhabitated between the Paleolithic and the Iron Age and used also as stations on the trans-Apennine routes \cite{men76, gia96, pal63, zec72a, zec72b}.

@@

We have little evidence of roman settlements in the mountainous hinterland, and the scarce archaeological traces are concentrated in the cave of Buca La Piella, investigated in 1975 by the Centre for Archaeological Studies of Lucca; it has two entrances joined by a walkable tunnel and rooms of discrete dimensions that overlook the outside. In addition to numerous faunal remains and fragments of locally produced common pottery, twenty bronze coins belonging to the 3rd century AD, bronze and lead objects were found \cite{gia96, men81, cia03}.

In Longobard and Carolingian times, Lucca's Val di Lima was one of the three administrative districts into which the mountains were divided and was called “fines Contronenses” \cite{qui02, cia06, cia11}. The only find from this period is the Grotta di Arzale, a rock shelter that opens up north-west of Casoli \cite{gia96}.

The first attestation of the settlements of Casoli and Lacu (Lake) dates back to the 10th century \cite{bar41, gia96}. Most documents of this period show the fractioning and alienation of ecclesiastical heritage in favour of the city aristocracy, securing an accumulation of funds and power that was to form the basis of the subsequent domains \cite{qui02, gia96, for12, for15}.

Written sources mention a castle in Casoli from 1180, but its foundation must be earlier. Between the 13th and 15th centuries, the fortification was at the centre, along with the other castles of the Val di Lima, of clashes between Lucca, Pisa and Florence, with alternating fortunes, as it was a strategic border area for the power of Lucca. Today, very little is preserved and the area has been partly inhabited \cite{gia96, for12, rom16}.

The same document from 1180 mentions the church of San Donato, located in the main square of the village of Casoli. The current structure dates back to the 12th-13th centuries, with subsequent renovations. Initially dedicated only to San Donato, it acquired the double dedication after the abandonment of the church of Sant'Andrea de Lacu. The latter is located on a plateau to the east of Lake Casoli and is attested from 1260 \cite{ber18, gia96, cap17}, but already in 15th century we learn of its state of abandonment \cite{gia96, con12}. This was probably due to the depopulation of the lake area and the simultaneous strengthening of the town of Casoli, which was fortified and better protected during this unstable phase.

Today, the romanesque church of Sant'Andrea de Lacu is in a state of decay, with a rectangular plan, ending on the east side with a semi-circular apse. Inside, there is a reused element of the previous building, testifying to an older origin, probably contemporary with the ancient settlement of Lacu (Lake) mentioned in sources from the 10th century and no longer visible today.

In the early modern age, villages of the area experienced a progressive architectural renewal, which still largely characterizes the settlements today. Within Casoli, a series of residential buildings with imposing dated portals are preserved, some with courtyards, and mansions that denote a discrete deployment of resources by wealthy social classes. Also dating from the modern era is the Madonna di Castello Oratory, characterised by an entrance eclosure and located along the cobbled road that traces the ancient route between the village and the summit area of the medieval castle \cite{gia96}.

Along the road that led toward Lucchio crossing the area of Casoli Lake there are the chapel of Madonna di Col di Piano and two “metati” referable to the contemporary age, used for drying chestnuts. Referred to as the "bread tree," chestnut fruits and the flour derived from them were a staple of the mountain diet until the mid-1900s \cite{buc92, puc10}.

@@

% TODO: \usepackage{graphicx} required
\begin{figure}
	\centering
	\includegraphics[width=\linewidth]{figure/mappa}
	\caption[Map with QR-tag positions]{Map with QR-tag positions}
	\label{fig:mappa}
\end{figure}
